Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-09-30 paper/manuscript.tex
Tweaks from @megosato

\paragraph{The choice of solvent force field may impact the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.

\paragraph{Charge models are non-trivial}
Modeling the partial charges of the solute and solvent molecules is also highly important. In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}. This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}. In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility. Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.

@@

\paragraph{Conformational sampling can also solvation and binding free energies}
While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial. Even subtle degrees of freedom such as orientations of the hydroxyl group (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging. In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.

>>>>

\paragraph{The choice of solvent force field may impact the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced different results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.

@@

\paragraph{Conformational sampling challenges can affect both solvation and binding free energies}
While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial. Even subtle degrees of freedom such as orientations of the hydroxyl group (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging. In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.
